---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---


## Loading and preprocessing the data

First, we will load the activity monitoring data and try to understand what it contains, as well as loading any required libraries

```{r dataload}

require(lattice)

# Unzip data
datazip<-("./activity.zip")

unzip(datazip)

# Load data
activity<-read.csv('activity.csv')

# Take a quick look at the data
head(activity)
str(activity)
summary(activity)
```

We can see that the data contains three variables. These are:

* steps
* date
* interval

## What is mean total number of steps taken per day?
To find the mean total number of steps per day the data first needs to be aggregated by date. A histogram can then be created and mean and median values displayed 

```{r meansteps}
dailysteps<-aggregate(steps~date,activity,sum)

hist(dailysteps$steps)

mean(dailysteps$steps)
median(dailysteps$steps)
```

## What is the average daily activity pattern?
To find the average daily activity pattern the data can be summed by interval and plotted, and the interval with the largest average number of steps

```{r dailyactivity}

averageday<-aggregate(steps~interval,activity,mean)

plot(averageday,type="l")

#interval with highest average number of steps
averageday[which.max(averageday$steps),]
```

## Imputing missing values
Number of NA values is calculated and the values are replaced using the mean of the average daily steps for that interval.

```{r missing}

sum(is.na(activity$steps))

# Make separate dataframe of N/As
nas<-activity[is.na(activity$steps),]
nas$steps<-NULL


# and replace the values from the dailysteps calculated
replacenas<-merge(nas,averageday, by.y="interval")

# Make new dataframe to be completed
activitycomplete<-activity

activitycomplete$steps[is.na(activitycomplete$steps)] <- replacenas$steps[match(activitycomplete$interval[is.na(activitycomplete$steps)],replacenas$interval)]

# Draw histogram
hist(activitycomplete$steps)

# Calculate mean and median
mean(activitycomplete$steps)
median(activitycomplete$steps)

```


## Are there differences in activity patterns between weekdays and weekends?

The weekday

```{r weekday}

#convert activity to date and add day column
activitycomplete$date<-as.Date(activitycomplete$date)
activitycomplete$day<-weekdays(activitycomplete$date)

#Create separate datasets for weekdays and weekends
weekend<-subset(activitycomplete,day!="Saturday"|day!="Sunday")
weekday<-subset(activitycomplete,day!="Saturday"&day!="Sunday")

#insert the weekend toggle
weekend$weekend<-"Weekend"
weekday$weekend<-"Weekday"

#rbind and aggregate the data
activitycomplete1<-rbind(weekend,weekday)
activitycomplete1<-aggregate(steps~interval+weekend,activitycomplete1,mean)

# Plot to compare the two

xyplot(steps~interval|factor(weekend),data=activitycomplete1,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))
```

